---
title: '30June16 explant sups plate 3: MIG, RANTES, IL1ra'
author: "Claire Levy"
date: "July 6, 2016"
output: github_document
---
This is an analysis of data from running our MSD plate 3 on some of the HSV2 infected explant sup samples for Herpes P01. See "explant_and_HVE_cell_sup_MSD_README.txt"" for more on the different plates/analyte panels. This plate has the following analytes measured in each well

Plate 3: 96 well 4 spot prototype Human Triplex

	Il-1ra
	MIG (CXCL9)
	RANTES (CCL5)
	


```{r setup, echo=FALSE, warning=FALSE, message= FALSE}
library(dplyr)
library(ggplot2)
library(pander)
library(reshape2)
library(stringr)
library(swarm)





#The MSD machine gives the data in an excel or text format. I have .csv file with the results. 


#read in the .csv raw results file
#../ to go up one level and then back down into a different folder. I exported this file from the DWB experiment called "Hladik plate 3 30 June 16 corrected layout". I imported that file from the folder location where FC saved the DWB experiment with the corrected layout. That file is here:

#J:\MacLabUsers\HLADIK SHARED\Projects\Herpes study\MSD\CVL Test Samples\Raw Data from MSD\Test Run

#I appended the name with "corrected layout" because I imported the wrong file at first and you can't delete them from the software.


dat<- read.csv("../raw_data/30June16_plate3_CVL_and_explant_sup_CLexport.csv", skip = 1)




### Formatting note: In the data table, there is both " T3" and "T3" so I am fixing it here.

dat$Sample<- str_replace(dat$Sample, " T3", "T3")


#### NOTE ####
# This file also contains data from TEST CVL samples from the hvtn clinic that we were running to test the analytes before doing the actual trial samples from Kenya.


#make a column identifying what group the sample belongs to, standards or explant sup or CVL sample so I can filter out what I want.





dat<- dat %>%
  mutate(Type =
 ifelse(str_detect(Sample,"^3")==TRUE, "explant_sup",
        ifelse(str_detect(Sample, "R|V")==TRUE,
                          "CVL","Standard")))



## Also, we included the diluted stock calibrator and the blended working stock on the plate when we shouldn't have. So we excluded those samples in the Discovery workbench software, which re-sets the curve fit based on the remaining samples. When you export the data table, it still includes those EXCLUDED samples. I am removing them here.


dat <- dat %>%
  filter(Sample != "Diluted Stock Calibrator" & Sample != "Blended Working Stock")

```




```{r standards, echo=FALSE, warning=FALSE, message= FALSE}


#filtering out the standards
standards <- dat %>%
  filter(Type == "Standard")
  

#which standards are in the detection range
standardsInRange <-standards%>%
  filter(Detection.Range == "In Detection Range")


#which were not in the detection range?

standardsOutOfRange<-standards%>%
  filter(Detection.Range != "In Detection Range")%>%
  arrange(Type, Assay, Sample)%>%
  select(Sample, Assay, Detection.Range)



```

Of the `r nrow(standards)` standards that we measured, `r nrow(standardsInRange)` are in the detection range. The following standards were out of the detection range:
`r pander(standardsOutOfRange)`




Here is a plot of the % recovery (extrapolated concentration/ known concentration x 100) for the standards that were either within or above the detection range, with lines at the "good recovery" limits of 80 and 120% recovery.


```{r percent recovery, echo=FALSE, warning=FALSE, message= FALSE}

stdFit<- standards %>% 
  arrange(Sample,Assay)%>%
  filter(Detection.Range != "Below Fit Curve Range")%>%
  select(Type, Sample, Assay,Detection.Range, X..Recovery,Fit.Statistic..RSquared, Concentration, Calc..Concentration)


ggplot(stdFit, aes(x = Sample, y = X..Recovery))+
  geom_point(aes(color = Sample, shape = Detection.Range ),size = 3)+
  scale_shape_manual(values = c("Above Fit Curve Range" = 1, "In Detection Range" = 16))+
  facet_wrap(~Assay)+
  theme(axis.text.x = element_text(angle=25))+
  scale_y_continuous(limits = c(75, 200))+
  geom_abline(slope = 0, intercept = c(80, 120))+
  labs(y = "Percent Recovery")+
  theme(axis.text.x = element_text(size = rel(0.8)))
  
```

The R-squared values for goodness-of-fit were good for all assays:

```{r r-squared, echo=FALSE, warning=FALSE, message= FALSE}

rSquared<-stdFit%>%
  group_by(Assay)%>%
  summarize(R_Squared = mean(Fit.Statistic..RSquared))

pander(rSquared)
```


```{r standard curves, echo=FALSE, warning=FALSE, message= FALSE}
#according to MSD instructions, the lower detection limit for the standards (in signal units) is the signal for the lower calibrator + (statistical factor * multiplier)


# upper limit is calculated this way: upper plateau of standard curve - (multiplier x statistical factor). The upper plateau value is in the "Algorithm Parameter Calc Top" column.

#the statistical factor used can be found in the "Detection limits bottom of range type" column and the multiplier can  be found in the "Detection limits bottom of range parameter" column


#for the 4PL curve, std deviation is the statistical factor used for both limits.





# LowerLimits <-standards %>%
#   group_by(Assay)%>%
#   filter(Sample == "STD-08")%>%
#   summarize(LowerLimit = unique(Mean + (Std..Deviation * Detection.Limits..Bottom.of.Range.Parameter)))
# 
# UpperLimits<- standards %>%
#   group_by(Assay)%>%
#   summarise(UpperLimit = unique(Algorithm.Parameter..Calc..Top - (Std..Deviation * Detection.Limits..Top.of.Range.Parameter)))
# 
# standards<-merge(standards,LowerLimits, by = "Assay")
# 
# standards<- merge(standards, UpperLimits, by = "Assay")
# 
# 
# 
# MIGstandards<- filter(standards, Assay == "MIG")
# 
# MIGstandards%>%
#   ggplot(aes(x = Concentration, y = Signal))+
#   geom_point()+
#   scale_y_log10()+
#   scale_x_log10()+
#   geom_hline(yintercept = MIGstandards$UpperLimit[1])+
#   geom_hline(yintercept = MIGstandards$LowerLimit[1])+
#   ggtitle("MIG")
#   
# 
# RANTESstandards<- filter(standards, Assay == "RANTES")
# 
# RANTESstandards%>%
#   filter(Assay == "RANTES")%>%
#   ggplot(aes(x = Concentration, y = Signal))+
#   geom_point()+
#   scale_y_log10()+
#   scale_x_log10()+
#   geom_hline(yintercept = RANTESstandards$UpperLimit[1])+
#   geom_hline(yintercept = RANTESstandards$LowerLimit[1])+
#   ggtitle("RANTES")
# 
# IL1RAstandards<-filter(standards, Assay == "IL1-RA")
# 
# IL1RAstandards%>%
#   filter(Assay == "IL1-RA")%>%
#   ggplot(aes(x = Concentration, y = Signal))+
#   geom_point()+
#   scale_y_log10()+
#   scale_x_log10()+
#   geom_hline(yintercept = IL1RAstandards$UpperLimit[1])+
#   geom_hline(yintercept = IL1RAstandards$LowerLimit[1])+
#   ggtitle("IL1-RA")

```




```{r explant data cleanup, echo=FALSE, warning=FALSE, message= FALSE}

#filtering for the samples I want (just want explant sups, not CVLS or standards)

explantDat<-dat%>%
  filter(Type == "explant_sup")

#sanity check: I put 56 samples on the plate and there are 3 analytes so we should have 56 x 3 = 168 samples. 

#We didn't have enough sample to do two replicates for 324-T2-M so we didn't add anything to well F10. We did read the well though, so there is data for it, but I want to exclude it because there wasn't sample there


          




explantDat <- explantDat %>%
  filter(Well != "F10") #removes 3 data points, 1 for each analyte so 165 left

#which explant samples are in the detection range
explantDatInRange <-explantDat%>%
  filter(Detection.Range == "In Detection Range")


#which were not in the detection range?

explantDatOutOfRange<-explantDat%>%
  filter(Detection.Range != "In Detection Range")%>%
  arrange(Type, Assay, Sample)%>%
  select(Type,Sample, Assay, Detection.Range)

```



```{r adding ids and avg reps, echo=FALSE, warning=FALSE, message= FALSE}

# I want to be able to group the data by timepoint and donor and virus so I need to separate the sample name and make new columns identifying them that way


donorTimeVirus<-colsplit(explantDat$Sample,"-", c("Donor","Timepoint","Virus") )

#bind the split col back to explantDat

explantDat<-cbind(donorTimeVirus, explantDat)


#I should only have 1 replicate for 324-T2-M. All other samples have 2 reps. 


#There is one sample where the concentration was below the fit curve range and the concentration is given as "NaN". This is because values can't be extrapolated below the curve. I am removing these data points.


explantDat <- explantDat %>%
  filter(Detection.Range != "Below Fit Curve Range")

explantDat$Donor <- as.factor(explantDat$Donor)

explantDat$Calc..Concentration <-as.numeric(explantDat$Calc..Concentration)

explantDat$Type <- as.factor(explantDat$Type)

explantDat$Virus <- as.factor(explantDat$Virus)

explantDat$Virus<-factor(explantDat$Virus, levels = c("M", "V1","V2","M 1:50", "V2 1:50"))

explantDat$Timepoint<-as.factor(explantDat$Timepoint)


#Averaging the replicates. 
#Sanity check: 
#For 27/28 samples I have 2 replicates = 54 
# For 1/28 samples I have 1 replicate = 1
#If I average those samples, I get 54/2 + 1 = 28 avg values
# I have data for 3 analytes so I should have 28 * 3 average values.


# AvgConcExplantDat <- explantDat %>%
#   group_by(Sample, Assay, Donor, Timepoint,Virus, Detection.Range)%>%
#   summarize(avgConc = mean(Calc..Concentration))
```
## Caveats

Some data points were above the range of the curve fit. The MSD software can extrapolate values for these if the curve is linear at the top. Values cannot be extrapolated when the data was below the fit curve range, so those are not shown in the following plots. 





```{r all analytes, echo=FALSE, warning=FALSE, message= FALSE}


ggplot(explantDat, aes(x = Virus, y = Calc..Concentration))+
  
  geom_point(aes(color = Assay, shape = Detection.Range ),size = 1.5)+
  
  scale_shape_manual(values = c("Above Fit Curve Range" = 1, "Below Fit Curve Range" = 1, "In Detection Range" = 16))+
  
  scale_y_log10()+
  
  labs( y = "Concentration")+
  
  facet_wrap(Timepoint~Donor, scales = "free_x")+
  
  theme(axis.text.x = element_text(angle = 25, size = 9))+

  ggtitle("Data from all analytes")


```


When we analyzed the explant microarray data, we saw the most obvious differences in gene expression were in V2T3 for donors 323,317,324,318. The only sample that was in this subset that we ran on this plate was 323.


```{r subset, echo=FALSE, warning=FALSE, message= FALSE}
subset <- explantDat %>%
  filter(Donor!="319" & Donor != "319" & Donor != "327" & Donor!= "326", Timepoint == "T3")



ggplot(subset, aes(x = Virus, y = Calc..Concentration))+
  
  geom_point(aes(color = Assay, shape = Detection.Range ),size = 1.5)+
  
  scale_shape_manual(values = c("Above Fit Curve Range" = 1, "Below Fit Curve Range" = 1, "In Detection Range" = 16))+
  
  scale_y_log10()+
  
  labs( y = "Concentration")+
  
  facet_wrap(Timepoint~Donor, scales = "free_x")+
  
  theme(axis.text.x = element_text(angle = 25, size = 9))

```


Data separated by analyte:


```{r MIG, echo=FALSE, warning=FALSE, message= FALSE}

#I want to make separate plots for each analyte, so I'll split it into different dfs here:


#MIG

MIGexplantDat<- explantDat %>%
  filter(Assay == "MIG")

ggplot(data = MIGexplantDat, aes(x = Virus, y = Calc..Concentration))+
  
  geom_point(aes(color = Donor, shape = Detection.Range ),size = 1, alpha = 0.5, position = position_swarm(0.5) )+
  
  stat_summary(aes(group = Donor, color = Donor),fun.y = mean, geom = "point", size = 2)+
  
  stat_summary(aes(group = Donor, color = Donor), fun.y =mean, geom = "line")+

  
  scale_shape_manual(values = c("Above Fit Curve Range" = 1, "Below Fit Curve Range" = 1, "In Detection Range" = 16))+
  
  scale_y_log10()+
  
  labs( y = "Concentration")+
  
  facet_wrap(Timepoint ~ Donor, scales = "free_x")+
  
  theme(axis.text.x = element_text(angle = 25))+
  ggtitle("MIG data")



```

```{r RANTES, echo=FALSE, warning=FALSE, message= FALSE}
#RANTES

RANTESexplantDat<- explantDat %>%
  filter(Assay == "RANTES")

ggplot(data = RANTESexplantDat, aes(x = Virus, y = Calc..Concentration))+
  
  geom_point(aes(color = Donor, shape = Detection.Range ),size = 1, alpha = 0.5, position = position_swarm(0.5) )+
  
  stat_summary(aes(group = Donor, color = Donor),fun.y = mean, geom = "point", size = 2)+
  
  stat_summary(aes(group = Donor, color = Donor), fun.y =mean, geom = "line")+

  
  scale_shape_manual(values = c("Above Fit Curve Range" = 1, "Below Fit Curve Range" = 1, "In Detection Range" = 16))+
  
  scale_y_log10()+
  
  labs( y = "Concentration")+
  
  facet_wrap(Timepoint ~ Donor, scales = "free_x")+
  
  theme(axis.text.x = element_text(angle = 25))+
  ggtitle("RANTES data")


```




```{r IL1-ra, echo=FALSE, warning=FALSE, message= FALSE}


IL1RAexplantDat<- explantDat %>%
  filter(Assay == "IL1-RA")

ggplot(data = IL1RAexplantDat, aes(x = Virus, y = Calc..Concentration))+
  
  geom_point(aes(color = Donor, shape = Detection.Range ),size = 1, alpha = 0.5, position = position_swarm(0.5) )+
  
  stat_summary(aes(group = Donor, color = Donor),fun.y = mean, geom = "point", size = 2)+
  
  stat_summary(aes(group = Donor, color = Donor), fun.y =mean, geom = "line")+

  
  scale_shape_manual(values = c("Above Fit Curve Range" = 1, "Below Fit Curve Range" = 1, "In Detection Range" = 16))+
  
  scale_y_log10()+
  
  labs( y = "Concentration")+
  
  facet_wrap(Timepoint ~ Donor, scales = "free_x")+
  
  theme(axis.text.x = element_text(angle = 25))+
  ggtitle("IL1-RA data")


```

